Type-Safe Modular Parsing

### 2016-10-18 paper/paper.tex
More text.
Text

% "define" code highlights for Java and Scala
\lstdefinelanguage{JavaScala}{
  morekeywords={public,int,interface,implements,default,
    abstract,case,catch,class,def,static,%
    do,else,extends,false,final,finally,%
    for,if,implicit,import,match,mixin,%
    new,null,object,override,package,%
    private,protected,requires,return,sealed,%
    super,this,throw,trait,true,try,%
    type,var,while,yield},
  otherkeywords={=>,<-,<\%,<:,>:,\#,@},
  sensitive=true,
  morecomment=[l]{//},
  morecomment=[n]{/*}{*/},
  morestring=[b]",
  morestring=[b]',
  morestring=[b]"""
}

@@

\input{sections/Library/OpenRecursion.tex}
\input{sections/Library/ObjectAlgebras.tex}
\input{sections/Library/FirstTry.tex}
\input{sections/Library/ParserCombinators.tex}
\input{sections/Library/DifferentSyntax.tex}

% more: two pretty-printers OldPretty and NewPretty, they are separated to show modularity.
% subsection: modify the code using List[E] and type bounds.
% more subsections: lexing, traits, operations, language composition (more extensibility)

>>>>

% "define" code highlights for Java and Scala
\lstdefinelanguage{JavaScala}{
  morekeywords={public,int,interface,implements,default,
    abstract,case,catch,class,def,static,%
    do,else,extends,false,final,finally,%
    for,if,implicit,import,match,mixin,%
    new,null,object,override,package,%
    private,protected,requires,return,sealed,%
    super,this,throw,trait,true,try,%
    type,var,while,yield,with,val},
  otherkeywords={=>,<-,<\%,<:,>:,\#,@},
  sensitive=true,
  morecomment=[l]{//},
  morecomment=[n]{/*}{*/},
  morestring=[b]",
  morestring=[b]',
  morestring=[b]"""
}

@@

\input{sections/Library/OpenRecursion.tex}
\input{sections/Library/ObjectAlgebras.tex}
\input{sections/Library/ParsingLibrary.tex}
\input{sections/Library/FirstTry.tex} % mention again | and ||| with concrete examples. (pApp | pVar)
\input{sections/Library/DifferentSyntax.tex}

% Object Algebras
% type synonyms, Parser[Expr], Fix[..]
% more: two pretty-printers OldPretty and NewPretty, they are separated to show modularity.
% subsection: modify the code using List[E] and type bounds.
% more subsections: lexing, traits, operations, language composition (more extensibility)

### 2016-10-18 paper/sections/CaseStudy.tex
Text

To demonstrate the utility of our modular parsing library, we implemented parsers of the first 18 calculi from the book \textit{Types and Programming Languages}. In this book, it introduces several calculi from simple ones to complex ones by gradually adding new features to syntax. It is suitable for our case study for mainly two reason. First, the calculi are arguably practical as we want to show our library could be used in real world development. One of the most complex calculi in our case study is System F (with several extensions such as records and variants), which could be the foundation of practical functional programming languages. Second, the evolution of calculi in the book reveals the advantages of modular representation of abstract syntax and modular parsing, which is the key functionality of our library. By decomposing those calculi into smaller components and reusing the common ones, we can obtain considerably code reuse in parsing with little performance penalty, as shown later.

For better modularity and eliminate code duplication, we extract 21 reusable \textit{components} from syntax of all the calculi. Each component, which may contain several syntactical structures, is a reusable fragment representing a certain feature in the syntax. For example, \lstinline{VarApp} component represents the variable case and application case of expressions. In our case study, each component is represented by a Scala object which has three parts. \lstinline{Alg} is an object algebra interface for the abstract syntax. \lstinline{Print} is an object algebra implements that interface \lstinline{Alg}, for the pretty printing operation. \lstinline{Parser} is the modular parser for this piece of syntax.

@@

Then each calculus could be composed directly from components and other calculi, if it has those common structures in the syntax. For example, the calculus \lstinline{Untyped}, representing the famous untyped lambda calculus, can be constructed from components \lstinline{VarApp} and \lstinline{UntypedAbs}.

>>>>

To demonstrate the utility of our modular parsing library, we implemented parsers of the first 18 calculi from book \textit{Types and Programming Languages}. In the book it introduces several calculi from simple to complex, by gradually adding new features to syntax. It is suitable for our case study for mainly two reason. First, the calculi are arguably practical as we want to show our library could be used in real world development. One of the most complex calculi in our case study is System F (with several extensions such as records and variants), which could be the foundation of practical functional programming languages. Second, the evolution of calculi in the book reveals the advantages of modular representation of abstract syntax and modular parsing, which is the key functionality of our library. By decomposing those calculi into smaller components and reusing the common ones, we can obtain considerably code reuse in parsing with little performance penalty, as shown later.

For better modularity and eliminate code duplication, we extract 21 reusable \textit{components} from syntax of all the calculi. Each component, which may contain several syntactical structures, is a reusable fragment representing a certain feature in the syntax. For example, \lstinline{VarApp} component represents the variable case and application case of expressions. In our case study, each component is represented by a Scala object which has three parts. \lstinline{Alg} is an object algebra interface for the abstract syntax. \lstinline{Print} is an object algebra implements that interface \lstinline{Alg}, for the pretty printing operation. It is not necessary for parsing, but we include it here for demonstration. \lstinline{Parser} is the modular parser for this piece of syntax.

@@

We use some naming conventions in our code, \lstinline{E} represents expressions, \lstinline{T} represents types and \lstinline{K} represents kinds. In this component \lstinline{VarApp} we only have expressions. The \lstinline{Alg} and \lstinline{Print} are standard object algebra operations, thus we will focus on the modular parser \lstinline{Parser}. Each parser carries lexical information and parsing functions. The lexical information includes reserved words, delimiters, etc. Parsing functions are similar with we described before, except that we use record types under some subtyping constraints here for the explicit self-reference. The record type contains a parser for expressions \lstinline{pE}, a parser for types \lstinline{pT} and a parser for kinds \lstinline{pK}, if the corresponding one is needed when parsing the syntax of this component. Here the type of self-reference, namely \lstinline{F}, is only required to have a parser of expressions \lstinline{pE}. Then we can build the parsing function \lstinline{pVarAppE}, the suffix \lstinline{E} also denotes that it will parse an expression.

Each calculus could be composed directly from components and other calculi, if it has those common structures in the syntax. For example, the calculus \lstinline{Untyped}, representing the famous untyped lambda calculus, can be constructed from components \lstinline{VarApp} and \lstinline{UntypedAbs}.
